{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-11-28 thesis/related.tex
dave's (easy) revisions to related and future

\section{Systematic Concurrency Testing}

@@

Systematic concurrency testing dates back to Verisoft \cite{verisoft},
the 1997 tool which first attempted to exhaustively explore the possibile ways to interleave threads.
Since then, researchers have built many tools along the same lines to test many kinds of programs.
One of the best-known SCTs is Microsoft Research's CHESS \cite{chess},
a checker for userspace C++ programs which preempts on synchronization APIs by default,
supporting compiler instrumentation to preempt on memory accesses as well,
and which pioneered the ICB search strategy discussed below.

Many checkers exist which target programs written for various different types of
concurrent execution and/or programming environments.
MaceMC \cite{macemc}, MoDist \cite{modist}, SAMC \cite{samc}, ETA \cite{dbug-retreat}, and Concuerror \cite{concuerror},
focus on distributed systems, where concurrent events are limited to message-passing and may span across multiple machines.
R4 \cite{r4} and EventRacer \cite{eventracer} check event-driven concurrent programs typical in mobile applications.
Like Landslide, SimTester \cite{simtester} is a Simics~\cite{simics}-based tool for kernel-level code,
although it focuses on interrupt nondeterminism for testing device drivers,
and is limited to injecting at most one interrupt per test run (as if under ICB with a bound of 1).
%
%Other checkers target specific programming languages' concurrency models and/or thread communication APIs.
dBug \cite{dbug-ssv}, another CMU original similar to CHESS,
tests natively-executing programs
using a dynamic library preload to insert preemption points at pthread and MPI interface boundaries.
Inspect \cite{inspect} uses a static alias analysis to instrument and preempt all memory accesses to potentially-shared data
at compile time, in addition to common synchronization APIs.
RacePRO~\cite{racepro} targets multi-process programs using system calls such as the filesystem API as preemption points
to find bugs which can corrupt persistent system resources.
SPIN \cite{spin} tests algorithms defined in the PROMELA domain-specific language,
instruments every memory access,
uses explicit state tracking rather than the stateless approach (\sect{\ref{sec:overview-stateless}}),
and specializes in verifying synchronization primitives such as RCU \cite{rcu}.
TLC \cite{tlc} checks formal models of concurrent program behaviour
written in the specification language TLA+ \cite{tlaplus},
and is arguably one of the only true concurrency {\em model checkers}
as it checks specifications separate from the programs themselves
%using explicit state tracking,
rather than attempting to exhaustively exercise every thread interleaving directly.
%
D\'{e}j\`{a} Fu \cite{dejafu} is a SCT
for the Haskell language,
whose strong type system guarantees that thread communication be confined to trusted, type-safe APIs.
%to a trusted API that implements internal synchronization to preserve type safety.
%Supporting both abstraction reduction and STM,
It instruments these interfaces (STM among them)
to check for deadlocks or nondeterministic behaviour in general,
which either may arise despite the static no-data-race guarantee.

The problem of relaxed memory nondeterminism alone has inspired the creation of several new SCTs in the past few years.
Relacy \cite{relacy}, a header-only C++ SCT library for checking synchronization primitives,
was the first to broach this field,
although requires custom annotations for non-atomic memory accesses
and
%(according to later citing papers) % cdschecker
does not fully model all possible relaxed memory behaviours.
%designed for verifying synchronization primitives,
CDSChecker \cite{cdschecker} extends DPOR with a {\em reads-from} relation
to capture most of the C++11 memory model's new behaviours.
Nidhugg \cite{nidhugg} is a SCT for TSO and PSO which instruments LLVM abstract assembly,
% i don't actually know what the difference is #overlyhonestmethods
although does not yet support the C++11 memory model.
%% actually it doesn't -- they say they take the Source DPOR worse version. im confus?
% and extends the Optimal DPOR algorithm (discussed below) to include store buffer nondeterminism.
rInspect \cite{tsopso}
%models TSO and PSO differently, using shadow threads, and
offers further heuristic state space reduction using buffer bounding (described below).
RCMC \cite{rcmc} models a ``repaired'' version of the C++11 memory model known as RC11 \cite{rc11},
and professes to achieve the best state space reduction to date.
These tools each use various heuristics to account for spin-wait loops,
ranging from delay bounding \cite{bpor} to a rigid rewrite rule,
and provide only limited support so far for read-modify-write atomics
(at best, supporting them by introducing % oops
some redundant exploration).
%
No relaxed-memory SCT has yet proposed a satisfactory model for the ``thin-air'' problem \cite{sully-thesis},
which can cause state space cycles in a way not yet well-understood and remains future work.
%and notably includes relaxed memory nondeterminism in its concurrency model.
They also identify all data races
% TODO check if ther'es not a better section reference
(under the C++ definition rather than \sect{\ref{sec:quicksand-soundness}}'s)
as bugs immediately, rather than checking them for benign or buggy outcomes.
All the tools in this paragraph are notably open-source -- an encouraging recent trend in the field.

@@

To date a number of techniques have been proposed to mitigate exponential explosion,
the Sisyphean rock of SCT.
The notion that some interleavings of concurrent threads could lead to indistinguishable program states and be therefore redundant,
known as {\em partial order reduction} (POR),
was first proposed in \cite{partial-model-checking}
and explored in detail in \cite{partial-order-methods}.
{\em Dynamic POR} (DPOR) was later developed in \cite{dpor},
proposing to track communication events between threads on-the-fly (i.e., dynamically)
rather than to rely on imprecise static alias analyses,
and is now considered the baseline for all subsequent state space reduction approaches in SCT.
That paper includes the {\em sleep sets} extension,
which Landslide includes in its implementation.
It is a {\em sound} reduction algorithm, meaning it will never fail to test a possible program behavior, despite skipping many execution sequences.
\sect{\ref{sec:landslide-dpor}} provides a detailed walk-through of how DPOR works,
as many of this thesis's contributions build directly upon it.
%
DPOR has since been extended in several ways to achieve further reduction
and to incorporate new concurrency models.
\revision{Distributed DPOR \cite{parallel-dpor}
allows the exploration to be parallelized,
with a minimum of overhead from redundant interleavings that would ordinarily be pruned in sequential DPOR.}
Optimal DPOR \cite{optimal-dpor} extends sleep sets into the more expressive {\em wakeup trees},
which provably tests exactly one interleaving from each equivalence class,
i.e., the optimal possible reduction,
at least under the memory independence definition of equivalence.
Extending the equivalence relation itself to capture not just memory {\em address} conflicts
but also the {\em values} read and written,
SATCheck \cite{satcheck} and Maximal Causality Reduction (MCR) \cite{mcr}
use an SMT solver \cite{z3} to identify additional pruning opportunities.
Implementing parallelization, wakeup trees, or SMT-driven exploration in Landslide is left to future work.

@@

Of course, no matter how optimal a sound reduction, there will always be programs too large to test.
To provide even partial results for state spaces that exceed the testing budget
(whether as predicted by automatic estimation \cite{estimation} or by a human's wild guess),
various heuristic exploration strategies have been proposed.
Preemption Sealing \cite{sealing} allows programmers to manually exclude preemption points
arising from trusted source code modules;
Landslide implements this as the {\tt without\_function} command (\sect{\ref{sec:landslide-pps}}).
Iterative Context Bounding (ICB) \cite{chess-icb} (\sect{\ref{sec:landslide-icb}})
orders the search space by increasing number of preemptions in each branch,
which is empirically more likely to expose bugs sooner should they exist;
BPOR \cite{bpor} extends DPOR to preserve soundness
thereunder. % what a word
Landslide implements ICB and BPOR for Chapter~\ref{chap:quicksand}'s control experiments,
although does not yet incorporate it into this thesis's own contributions
(as discussed in Chapter~\ref{chap:warpzone}).
Chapter~\ref{chap:quicksand}'s Quicksand algorithm is, effectively, another such heuristic search strategy,
focusing on preemption point subsets rather than context switch bounding.
\revision{Maple \cite{maple} proposed a heuristic metric for measuring the amount of ``concurrency coverage'',
analogous to line-by-line coverage for sequential programs,
and prioritizes testing interleavings which increase the coverage metric,
although it makes several limiting assumptions such as
% "vulnerability window", in its words
preemption locality
% this is like, TOTAL bs, completely results-oriented empiricism
% and at worst, intentionally blinding yourself to the actually difficult bugs
and independence of values read/written.}
DeMeter \cite{demeter} adapted abstraction reduction to distributed systems verification under the name Dynamic Interface Reduction,
while dBug \cite{dbug-phdthesis} applied abstraction reduction to synchronization primitives,
and I showed how it could be applied similarly to transactional memory in \sect{\ref{sec:tm-abstraction}}.
Each of these approaches is compatible (and indeed, throughout this thesis used often in concert)
with the sound reduction analyses listed above.
% mentioning PCT would disrupt the flow here and it's a dumb technique anyway so i won't bother

@@

Although many SCTs listed in the previous section are content to report any data races as outright bugs,
RacerX \cite{racerx} showed that tools must be careful not to overwhelm users with benign warnings
they don't care about fixing.
This has motivated replay analysis
to classify data-race candidates by their impact on program behaviour
by extending single-pass data-race analysis to many thread interleavings.
It was first introduced in \cite{recordreplaydrs},
which compares the program states immediately after the access pair for differences,
preferring still to err on the side of false positives (as different program states might not necessarily lead to a failure).
RaceFuzzer \cite{racefuzzer} avoids false positives by requiring an actual failure be exhibited, as Quicksand does,
although it uses random schedule fuzzing rather than systematic testing for its concurrency coverage.
%
Portend \cite{portend} is closest in spirit to Quicksand:
it tests alternate executions based on single-pass data-race candidates to classify them in a taxonomy of likely severity,
including non-failing races which nevertheless cause nondeterministic output
in addition to obvious failures.
However, it does not
test alternate interleavings in advance of knowing any specific data races,
% TODO more speicifc section refrance
which \sect{\ref{sec:quicksand-eval}} showed is necessary to find certain bugs.
Quicksand builds on Portend's approach by introducing a feedback loop between the data-race analysis and SCT,
which results in a stronger verification property
when the test can be fully completed (\sect{\ref{sec:quicksand-soundness}}).
Portend also uses symbolic execution to test input nondeterminism as well as schedule nondeterminism,
while Quicksand remains at the mercy of manual test case design.
Future work could incorporate Portend's taxonomy to better help the user understand any non-failing data races
when the test is too large to complete,
as well as its symbolic execution to help user-provided tests achieve better coverage automatically.

@@

To my knowledge, this is the first study of model checking in an educational setting,
although teaching concurrency is not itself an unstudied problem.
%% tf??
% Eytani et al.~\cite{towards-a-framework} present a promising framework for testing concurrent programs,
% which can incorporate model checking as well as static analysis, resource exhaustion, data-race analysis, and coverage analysis.
% However, it lacks an evaluation, and makes mention of its educational value only in its future work remarks.
%
%One recent study~
\cite{how-studence} surveyed how students think about
testing and debugging during a concurrent programming project,
finding that unguided, students often approach testing haphazardly,
not understanding the goal of good concurrency coverage,
and also had difficulty understanding single failing executions.
In fact, the study explicitly recommended tool support for testing many thread interleavings automatically (SCT)
and for execution traces to communicate sequences of important events (preemption traces),
which I dare say I have achieved in this thesis.
A more recent study \cite{novices-programmers}
examined in detail the students' thought process
during the diagnosis and fixing phases,
although its participants were drawn from novice-level programming classes,
and the experiment was set up with more elementary bugs like syntax and logic errors correspondingly.
Nevertheless, the authors recommended teaching debugging skills explicitly
via systematic exposure to different kinds of bugs,
which suggests future work for even advanced operating systems curricula
to offer a ``warm up'' Landslide assignment
(for example, the {\tt atomic\_*} tests from \sect{\ref{sec:education-pebbles-tests}})
that could ultimately lead to a higher solve rate on Landslide's bug reports during P2 (\sect{\ref{sec:education-eval-bugs-cmu}}).

@@

@@

Transactional memory (TM), first introduced in 1993 \cite{transactional-memory},
% TODO expand this to talk about some stm implementations?
has received renewed attention in recent years with the announcement of Intel's Haswell architecture \cite{htm-haswell},
which supports hardware transactions (HTM) using new x86 instructions.
Since then, many studies have evaluated the increased performance it offers over traditional locking and/or STM
\cite{htm-experience,htm-performance,tm-benchmark-cmu}.
%
HTM's performance comes at an increased cost in complexity to the programmer,
who must avoid system calls or transaction nesting, respect the CPU cache capacity,
and consider retry loops for spurious failure.
SI-TM \cite{si-tm} introduces techniques for reducing HTM's abort rates for performance's sake,
but without eliminating them altogether, any full verification must still consider them possible anywhere.
For programmers who wish to avoid such concerns,
the simpler STM programming model remains relevant.
One recent work \cite{hybrid-htm-stm} enhances STM transactions to nest with HTM ones,
while another \cite{stm-relaxed-memory} adds support for relaxed memory models.
Meanwhile, two recent papers \cite{relaxed-transactions-pldi,relaxed-transactions-popl}
have proposed formal models of HTM's execution semantics under relaxed memory likewise.
Such extensions come with the challenge of even more complicated behavioural semantics
for SCT to accurately model and verify in future work.

Testing approaches for transactional programs are sparsely represented in the literature so far.
Although several related works \cite{tm-correctness,tm-completeness,specifying-verifying-tm}
are building up to formal proofs of the correctness of underlying TM {\em implementations},
Landslide is the first I know of to verify client programs thereof.
McRT STM \cite{mc-tm-with-spin} uses SPIN \cite{spin} to model check an STM implementation
up to 2 threads running 1 transaction each with up to 3 memory accesses.
This kind of verification,
analogous to \sect{\ref{sec:education-pebbles-tests}}'s {\tt mutex\_test},
is an important stepping stone for trusting the results Landslide will provide.
STAMP \cite{stamp} is a benchmark suite transactional programs,
implemented using the OpenTM interface \cite{opentm},
used by many papers in the field to evaluate the performance of both STM and HTM implementations alike.
As discussed in \sect{\ref{sec:tm-eval-exp-setup}},
it focuses more on performance than on interesting concurrency properties.
% stamp bugfixes - https://github.com/mfs409/transmem/blob/master/benchmarks/stamp_c/VERSIONS.
% stamp need not be considered harmful put here
Even so, the more recent Stampede suite \cite{scalable-tm} argues that STAMP's benchmarks
were constructed under a programming model poorly-suited to fully take advantage of HTM's performance,
and that scalable HTM programs must
minimize incidental conflicts and handle aborts more flexibly than with blind retry loops.
The programming complexity needed to achieve these goals calls,
of course,
for correspondingly advanced verification approaches such as Landslide.
Finally, TxRace \cite{txrace}
tests non-transactional programs for data races
by inserting HTM calls via compiler instrumentation,
% they don't actually seem to check the failure reason lmao
% i'll phrase this in a slightly weasel way to point out that's what they *should* be doing
% and that in turn motivates the stuff i did
% even tho they use a diff way to filter them post hoc (well, they'd still need to, bc of false sharing)
relying on conflict aborts to point out access pairs that would be unsafe in the original program.
This citation arguably belongs in \sect{\ref{sec:related-data-race}} as well;
I include it here to highlight the importance of Landslide's
ability to distinguish different abort reasons (\sect{\ref{sec:txn-abort-codes}}).

@@

While C's extremely rudimentary type system allows the compiler to statically check programs
for properties such as not accidentally dereferencing raw integer values as if they were pointers,
more advanced programming languages may make guarantees about concurrent execution.
%
Erlang \cite{erlang}, an early concurrent functional language,
introduced the actor model for concurrency,
in which threads share no state and must communicate only by message-passing.
%One might say this statically guarantees freedom from data races;
%from another point of view, data races do not even exist as a concept in this programming model.
While this statically guarantees the absence of data races,
programs may still execute nondeterministically,
so concurrency bugs, especially deadlocks, are not ruled out.
Concuerror \cite{concuerror}, discussed above, is a SCT tool for Erlang programs.
%
Haskell \cite{haskell} offers a more sophisticated interface to concurrency:
threads may reference the same objects and even update shared references
using monads that encapsulate mutation,
but at the (garbage-collected) execution level all data is immutable once created,
which preserves type soundness and data-race freedom.
The aforementioned D\'{e}j\`{a} Fu \cite{dejafu} checks concurrent Haskell programs.
%
Rust \cite{rust-language}
presents a type system with more explicit memory management,
in-place mutation,
and mutable references
to appear familiar and approachable to those already versed in C++.
It proposes a borrow-check analysis to ensure memory and type safety
despite mutable references,
and a trait system to ensure no shared state between threads by default.
Its concurrency libraries then offer interfaces which relax this restriction,
allowing threads even to simultaneously reference shared mutable state,
using the type system to enforce sound locking discipline across such accesses,
again preserving type soundness and data-race freedom%
\footnote{The author themself contributed the original design for this latter feature.}.
I know of no existing SCT for Rust as of yet.
%
The Relaxed Memory Calculus \cite{sully-thesis}
proposes to extend C++ with annotations for weak memory atomics,
which allows for static formal analysis of memory access reorderings.
Although not ruling out data races,
this approach is an important step towards compilers
which can statically reason about program execution under more advanced concurrency models.
%
Finally, LVish \cite{lvish}
features a type system that enforces deterministic behaviour by construction,
using shared state called LVars which allow writes only in ways that update order is not observable.
This renders thread interleavings entirely irrelevant,
obviating any need for runtime verification,
but at the cost of a more restrictive programming model.

>>>>

\section{Stateless Model Checking}

@@

Stateless model checking dates back to Verisoft \cite{verisoft},
the 1997 tool which first attempted to exhaustively explore the \revisionminor{possible} ways to interleave threads.
Since then, researchers have built many tools along the same lines to test many kinds of programs.
One of the best-known MCs is Microsoft Research's CHESS \cite{chess},
a checker for userspace C++ programs which preempts on synchronization APIs by default,
supporting compiler instrumentation to preempt on memory accesses as well,
and which pioneered the ICB search strategy discussed below.

Many checkers exist which target programs written for various different types of
concurrent execution and/or programming environments.
MaceMC \cite{macemc}, MoDist \cite{modist}, SAMC \cite{samc}, ETA \cite{dbug-retreat}, and Concuerror \cite{concuerror},
focus on distributed systems, where concurrent events are limited to message-passing and may span across multiple machines.
R4 \cite{r4} and EventRacer \cite{eventracer} check event-driven concurrent programs typical in mobile applications.
Like Landslide, SimTester \cite{simtester} is a Simics~\cite{simics}-based tool for kernel-level code,
although it focuses on interrupt nondeterminism for testing device drivers,
and is limited to injecting at most one interrupt per test run (as if under ICB with a bound of~1).
%
%Other checkers target specific programming languages' concurrency models and/or thread communication APIs.
dBug \cite{dbug-ssv}, another CMU original similar to CHESS,
tests natively-executing programs
using a dynamic library preload to insert preemption points at pthread and MPI interface boundaries.
Inspect \cite{inspect} uses a static alias analysis to instrument and preempt all memory accesses to potentially-shared data
at compile time, in addition to common synchronization APIs.
RacePRO~\cite{racepro} targets multi-process programs using system calls such as the filesystem API as preemption points
to find bugs which can corrupt persistent system resources.
SPIN \cite{spin} tests algorithms defined in the PROMELA domain-specific language,
instruments every memory access,
uses explicit state tracking rather than the stateless approach (\sect{\ref{sec:overview-stateless}}),
and specializes in verifying synchronization primitives such as RCU \cite{rcu}.
TLC \cite{tlc} checks formal models of concurrent program behaviour
written in the specification language TLA+ \cite{tlaplus},
and is arguably one of the only true concurrency {\em model checkers}
as it checks specifications separate from the programs themselves
%using explicit state tracking,
rather than attempting to exhaustively exercise every thread interleaving directly.
%
D\'{e}j\`{a} Fu \cite{dejafu} is a model checker
for the Haskell language,
whose strong type system guarantees that thread communication be confined to trusted, type-safe APIs.
%to a trusted API that implements internal synchronization to preserve type safety.
%Supporting both abstraction reduction and STM,
It instruments these interfaces (STM among them)
to check for deadlocks or nondeterministic behaviour in general,
which either may arise despite the static no-data-race guarantee.

The problem of relaxed memory nondeterminism alone has inspired the creation of several new tools in the past few years.
Relacy \cite{relacy}, a header-only C++ model checking library for checking synchronization primitives,
was the first to broach this field,
although \revisionminor{it} requires custom annotations for non-atomic memory accesses
and
%(according to later citing papers) % cdschecker
does not fully model all possible relaxed memory behaviours.
%designed for verifying synchronization primitives,
CDSChecker \cite{cdschecker} extends DPOR with a {\em reads-from} relation
to capture most of the C++11 memory model's new behaviours.
Nidhugg \cite{nidhugg} is a checker for TSO and PSO which instruments LLVM abstract assembly,
% i don't actually know what the difference is #overlyhonestmethods
although does not yet support the C++11 memory model.
%% actually it doesn't -- they say they take the Source DPOR worse version. im confus?
% and extends the Optimal DPOR algorithm (discussed below) to include store buffer nondeterminism.
rInspect \cite{tsopso}
%models TSO and PSO differently, using shadow threads, and
offers further heuristic state space reduction using buffer bounding (described below).
RCMC \cite{rcmc} models a ``repaired'' version of the C++11 memory model known as RC11 \cite{rc11},
and professes to achieve the best state space reduction to date.
These tools each use various heuristics to account for spin-wait loops,
ranging from delay bounding \cite{bpor} to a rigid rewrite rule,
and provide only limited support so far for read-modify-write atomics
(at best, supporting them by introducing % oops
some redundant exploration).
%
No relaxed-memory MC has yet proposed a satisfactory model for the ``thin-air'' problem \cite{sully-thesis},
which can cause state space cycles in a way not yet well-understood and remains future work.
%and notably includes relaxed memory nondeterminism in its concurrency model.
They also identify all data races
% TODO check if ther'es not a better section reference
(under the C++ definition rather than \sect{\ref{sec:quicksand-soundness}}'s)
as bugs immediately, rather than checking them for benign or buggy outcomes.
All the tools in this paragraph are notably open-source -- an encouraging recent trend in the field.

@@

To date a number of techniques have been proposed to mitigate exponential explosion,
the Sisyphean rock of stateless MC.
The notion that some interleavings of concurrent threads could lead to indistinguishable program states and be therefore redundant,
known as {\em partial order reduction} (POR),
was first proposed in \cite{partial-model-checking}
and explored in detail in \cite{partial-order-methods}.
{\em Dynamic POR} (DPOR) was later developed in \cite{dpor},
proposing to track communication events between threads on-the-fly (i.e., dynamically)
rather than to rely on imprecise static alias analyses,
and is now considered the baseline for all subsequent state space reduction approaches in stateless MC.
That paper includes the {\em sleep sets} extension,
which Landslide includes in its implementation.
It is a {\em sound} reduction algorithm, meaning it will never fail to test a possible program behavior, despite skipping many execution sequences.
\sect{\ref{sec:landslide-dpor}} provides a detailed walk-through of how DPOR works,
as many of this thesis's contributions build directly upon it.
%
DPOR has since been extended in several ways to achieve further reduction
and to incorporate new concurrency models.
\revision{Distributed DPOR \cite{parallel-dpor}
allows the exploration to be parallelized,
with a minimum of overhead from redundant interleavings that would ordinarily be pruned in sequential DPOR.}
Optimal DPOR \cite{optimal-dpor} extends sleep sets into the more expressive {\em wakeup trees},
which provably tests exactly one interleaving from each equivalence class,
i.e., the optimal possible reduction,
at least under the memory independence definition of equivalence.
Extending the equivalence relation itself to capture not just memory {\em address} conflicts
but also the {\em values} read and written,
SATCheck \cite{satcheck} and Maximal Causality Reduction (MCR) \cite{mcr}
use an SMT solver \cite{z3} to identify additional pruning opportunities.
Implementing parallelization, wakeup trees, or SMT-driven exploration in Landslide is left to future work.

@@

Of course, no matter how \revisionminor{powerful} a sound reduction, there will always be programs too large to test.
To provide even partial results for state spaces that exceed the testing budget
(whether as predicted by automatic estimation \cite{estimation} or by a human's wild guess),
various heuristic exploration strategies have been proposed.
Preemption Sealing \cite{sealing} allows programmers to manually exclude preemption points
arising from trusted source code modules;
Landslide implements this as the {\tt without\_function} command (\sect{\ref{sec:landslide-pps}}).
Iterative Context Bounding (ICB) \cite{chess-icb} (\sect{\ref{sec:landslide-icb}})
orders the search space by increasing number of preemptions in each branch,
which is empirically more likely to expose bugs sooner should they exist;
BPOR \cite{bpor} extends DPOR to preserve soundness
thereunder. % what a word
Landslide implements ICB and BPOR for Chapter~\ref{chap:quicksand}'s control experiments,
although does not yet incorporate it into this thesis's own contributions
(as discussed in Chapter~\ref{chap:warpzone}).
Chapter~\ref{chap:quicksand}'s Quicksand algorithm is, effectively, another such heuristic search strategy,
focusing on preemption point subsets rather than context switch bounding.
\revision{Maple \cite{maple} proposed a heuristic metric for measuring the amount of ``concurrency coverage'',
analogous to line-by-line coverage for sequential programs,
and prioritizes testing interleavings which increase the coverage metric,
although it makes several limiting assumptions such as
% "vulnerability window", in its words
preemption locality
% this is like, TOTAL bs, completely results-oriented empiricism
% and at worst, intentionally blinding yourself to the actually difficult bugs
and independence of values read/written.}
DeMeter \cite{demeter} adapted abstraction reduction to distributed systems verification under the name Dynamic Interface Reduction,
while dBug \cite{dbug-phdthesis} applied abstraction reduction to synchronization primitives,
and I showed how it could be applied similarly to transactional memory in \sect{\ref{sec:tm-abstraction}}.
Each of these approaches is compatible (and indeed, throughout this thesis used often in concert)
with the sound reduction analyses listed above.
% mentioning PCT would disrupt the flow here and it's a dumb technique anyway so i won't bother

@@

Although many MCs listed in the previous section are content to report any data races as outright bugs,
RacerX \cite{racerx} showed that tools must be careful not to overwhelm users with benign warnings
they don't care about fixing.
This has motivated replay analysis
to classify data-race candidates by their impact on program behaviour
by extending single-pass data-race analysis to many thread interleavings.
It was first introduced in \cite{recordreplaydrs},
which compares the program states immediately after the access pair for differences,
preferring still to err on the side of false positives (as different program states might not necessarily lead to a failure).
RaceFuzzer \cite{racefuzzer} avoids false positives by requiring an actual failure be exhibited, as Quicksand does,
although it uses random schedule fuzzing rather than model checking for its concurrency coverage.
%
Portend \cite{portend} is closest in spirit to Quicksand:
it tests alternate executions based on single-pass data-race candidates to classify them in a taxonomy of likely severity,
including non-failing races which nevertheless cause nondeterministic output
in addition to obvious failures.
However, it does not
test alternate interleavings in advance of knowing any specific data races,
% TODO more speicifc section refrance
which \sect{\ref{sec:quicksand-eval}} showed is necessary to find certain bugs.
Quicksand builds on Portend's approach by introducing a feedback loop between the data-race analysis and model checking,
which results in a stronger verification property
when the test can be fully completed (\sect{\ref{sec:quicksand-soundness}}).
Portend also uses symbolic execution to test input nondeterminism as well as schedule nondeterminism,
while Quicksand remains at the mercy of manual test case design.
Future work could incorporate Portend's taxonomy to better help the user understand any non-failing data races
when the test is too large to complete,
as well as its symbolic execution to help user-provided tests achieve better coverage automatically.

@@

To my knowledge, this is the first study of model checking in an educational setting,
although teaching concurrency is not itself an unstudied problem.
%% tf??
% Eytani et al.~\cite{towards-a-framework} present a promising framework for testing concurrent programs,
% which can incorporate model checking as well as static analysis, resource exhaustion, data-race analysis, and coverage analysis.
% However, it lacks an evaluation, and makes mention of its educational value only in its future work remarks.
%
%One recent study~
\cite{how-studence} surveyed how students think about
testing and debugging during a concurrent programming project,
finding that unguided, students often approach testing haphazardly,
not understanding the goal of good concurrency coverage,
and also had difficulty understanding single failing executions.
In fact, the study explicitly recommended tool support for testing many thread interleavings automatically (model checking)
and for execution traces to communicate sequences of important events (preemption traces),
which I dare say I have achieved in this thesis.
A more recent study \cite{novices-programmers}
examined in detail the students' thought process
during the diagnosis and fixing phases,
although its participants were drawn from novice-level programming classes,
and the experiment was set up with more elementary bugs like syntax and logic errors correspondingly.
Nevertheless, the authors recommended teaching debugging skills explicitly
via systematic exposure to different kinds of bugs,
which suggests future work for even advanced operating systems curricula
to offer a ``warm up'' Landslide assignment
(for example, the {\tt atomic\_*} tests from \sect{\ref{sec:education-pebbles-tests}})
that could ultimately lead to a higher solve rate on Landslide's bug reports during P2 (\sect{\ref{sec:education-eval-bugs-cmu}}).

@@

\revision{The Deadlock Empire~\cite{deadlockempire} is an educational introduction to concurrent programming and debugging,
presented as an online puzzle-hunt of sorts, backed up by a fantasy-themed narrative.
Each puzzle requires the user to manually step through the execution of multiple threads
to find buggy interleavings.
%which in some cases requires the use of an ``expand'' button that
Though itself lacking in scientific analysis of its educational power,
it demonstrates the kind of user-friendly features and overall approach
that many students expressed desire for in their survey responses (\sect{\ref{sec:education-eval-survey}}).
Future work could extend Landslide's preemption traces to include similar interactivity.}

% TODO: more papers to look up:
% Tool Design and Student Testing Behavior in an Introductory Java Course
% Testing Strategies for the Automated Grading of Student Programs

@@

Transactional memory (TM), first introduced in 1993 \cite{transactional-memory},
% TODO expand this to talk about some stm implementations?
has received renewed attention in recent years with the announcement of Intel's Haswell architecture \cite{htm-haswell},
which supports hardware transactions (HTM) using new x86 instructions.
Since then, many studies have evaluated the increased performance it offers over traditional locking and/or STM
\cite{htm-experience,htm-performance,tm-benchmark-cmu}.
%
HTM's performance comes at an increased cost in complexity to the programmer,
who must avoid system calls or transaction nesting, respect the CPU cache capacity,
and consider retry loops for spurious failure.
SI-TM \cite{si-tm} introduces techniques for reducing HTM's abort rates for performance's sake,
but without eliminating them altogether, any full verification must still consider them possible anywhere.
For programmers who wish to avoid such concerns,
the simpler STM programming model remains relevant.
One recent work \cite{hybrid-htm-stm} enhances STM transactions to nest with HTM ones,
while another \cite{stm-relaxed-memory} adds support for relaxed memory models.
Meanwhile, two recent papers \cite{relaxed-transactions-pldi,relaxed-transactions-popl}
have proposed formal models of HTM's execution semantics under relaxed memory likewise.
Such extensions come with the challenge of even more complicated behavioural semantics
for MCs to accurately model and verify in future work.

Testing approaches for transactional programs are sparsely represented in the literature so far.
Although several related works \cite{tm-correctness,tm-completeness,specifying-verifying-tm}
are building up to formal proofs of the correctness of underlying TM {\em implementations},
Landslide is the first I know of to verify client programs thereof.
% TODO: citation as noun - global scan
\revision{\cite{mc-tm-with-spin} used SPIN \cite{spin} to model check McRT STM \cite{mcrt-stm},
an STM implementation built by Intel,
verifying}
up to 2 threads running 1 transaction each with up to 3 memory accesses.
This kind of verification,
analogous to \sect{\ref{sec:education-pebbles-tests}}'s {\tt mutex\_test},
is an important stepping stone for trusting the results Landslide will provide.
STAMP \cite{stamp} is a benchmark suite transactional programs,
implemented using the OpenTM interface \cite{opentm},
used by many papers in the field to evaluate the performance of both STM and HTM implementations alike.
As discussed in \sect{\ref{sec:tm-eval-exp-setup}},
it focuses more on performance than on interesting concurrency properties.
% stamp bugfixes - https://github.com/mfs409/transmem/blob/master/benchmarks/stamp_c/VERSIONS.
% stamp need not be considered harmful put here
Even so, the more recent Stampede suite \cite{scalable-tm} argues that STAMP's benchmarks
were constructed under a programming model poorly-suited to fully take advantage of HTM's performance,
and that scalable HTM programs must
minimize incidental conflicts and handle aborts more flexibly than with blind retry loops.
The programming complexity needed to achieve these goals calls,
of course,
for correspondingly advanced verification approaches such as Landslide.
Finally, TxRace \cite{txrace}
tests non-transactional programs for data races
by inserting HTM calls via compiler instrumentation,
% they don't actually seem to check the failure reason lmao
% i'll phrase this in a slightly weasel way to point out that's what they *should* be doing
% and that in turn motivates the stuff i did
% even tho they use a diff way to filter them post hoc (well, they'd still need to, bc of false sharing)
relying on conflict aborts to point out access pairs that would be unsafe in the original program.
This citation arguably belongs in \sect{\ref{sec:related-data-race}} as well;
I include it here to highlight the importance of Landslide's
ability to distinguish different abort reasons (\sect{\ref{sec:txn-abort-codes}}).

@@

While C's extremely rudimentary type system allows the compiler to statically check programs
for properties such as not accidentally dereferencing raw integer values as if they were pointers,
more advanced programming languages may make guarantees about concurrent execution.
%
\revision{Language-level concurrency abstractions were first introduced by the Communicating Sequential Processes (CSP)
model~\cite{communicating-sequential-processes},
which defines a concurrent program as a set of isolated processes
allowed to communicate only via blocking input/output commands.
Erlang \cite{erlang}, an early concurrent functional language,
introduced the actor model for concurrency based on the CSP model,
in which threads likewise} share no state and must communicate only by message-passing.
%One might say this statically guarantees freedom from data races;
%from another point of view, data races do not even exist as a concept in this programming model.
While this statically guarantees the absence of data races,
programs may still execute nondeterministically,
so concurrency bugs, especially deadlocks, are not ruled out.
Concuerror \cite{concuerror}, discussed above, is a MC tool for Erlang programs.
%
Haskell \cite{haskell} offers a more sophisticated interface to concurrency:
threads may reference the same objects and even update shared references
using monads that encapsulate mutation,
but at the (garbage-collected) execution level all data is immutable once created,
which preserves type soundness and data-race freedom.
The aforementioned D\'{e}j\`{a} Fu \cite{dejafu} checks concurrent Haskell programs.
%
Rust \cite{rust-language}
presents a type system with more explicit memory management,
in-place mutation,
and mutable references
to appear familiar and approachable to those already versed in C++.
It proposes a borrow-check analysis to ensure memory and type safety
despite mutable references,
and a trait system to ensure no shared state between threads by default.
Its concurrency libraries then offer interfaces which relax this restriction,
allowing threads even to simultaneously reference shared mutable state,
using the type system to enforce sound locking discipline across such accesses,
again preserving type soundness and data-race freedom.%
\footnote{The author themself contributed the original design for this latter feature.}
I know of no existing model checker for Rust as of yet.
%
The Relaxed Memory Calculus \cite{sully-thesis}
proposes to extend C++ with annotations for weak memory atomics,
which allows for static formal analysis of memory access reorderings.
Although not ruling out data races,
this approach is an important step towards compilers
which can statically reason about program execution under more advanced concurrency models.
%
Finally, LVish \cite{lvish}
features a type system that enforces deterministic behaviour by construction,
using shared state called LVars which allow writes only in ways that update order is not observable.
This renders thread interleavings entirely irrelevant,
obviating any need for runtime verification,
but at the cost of a more restrictive programming model.
